Predicting the effects of character displacement on food-web dynamics

### 2020-07-02 manuscript_ECD_model.Rmd
add author info, acknowledgements, unstack inline equations, and final edit from AE.

---
output:
  pdf_document:
    keep_tex: true
    toc: false
    number_sections: false
    df_print: paged
    fig_caption: true
    citation_package: natbib 
    #includes:
    #  after_body: Figures_Edited.tex
header-includes:
  - \usepackage{fullpage}
  - \linespread{1.7} 
  - \usepackage{lineno}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
bibliography: ReferencesAbbrev.bib
biblio-style: amnat
fontsize: 11pt
fontfamily: mathpazo
documentclass: article
linkcolor: black
urlcolor: black
citecolor: black
---

@@

<!-- remove for double-blind peer-review
\bigskip

@@

*Data accessibility*: No new data were used. All code used to reproduce this manuscript and simulations within are publically available on GitHub (https://github.com/mabarbour/ECD_model) and has been archived with Zenodo (https://zenodo.org/badge/latestdoi/79349192).

@@

where consumer $j$'s feeding rate on resource $i$ is influenced by the abundance of each resource; saturates as resource abundances increase (due to handling time $h_{ij}$); and consumer habitat preferences are modified by the relative abundance of resources, such that: $W_{ij}=\frac{w_{ij}R_i}{w_{1,j}R_1+w_{2,j}R_2}$.

Previous studies have analyzed the evolution of consumer attack rates in the first two foraging scenarios using an Adaptive Dynamics approach, with the general result being divergent character displacement [@Lawlor1976; @Abrams1986]. 
I also used an Adaptive Dynamics approach to analyze the evolution of consumer attack rates in the third foraging scenario, and I too observed divergent character displacement (detailed analysis given in Appendix S1).
I say consumers have undergone divergent character displacement if their evolved attack rates are more specialized when evolving with vs. without a competing consumer. 
Specialization of consumer $j$ on resource $1$ is measured as $\frac{a_{1,j}}{a_{1,j}+a_{2,j}}$, where a value of 0.5 is a complete generalist ($a_{1,j}=a_{2,j}$), and a value of 1 is a complete specialist ($a_{2,j}=0$).
Values less than 0.5 indicate specialization on the other resource. 
Since I did not observe convergent character displacement in any of the foraging scenarios I analyzed, I refer to divergent character displacement as simply (ecological) character displacement throughout the rest of the text.

@@

To gain insight to the eco-evolutionary feedback generated by character displacement, I conducted simulations using an Adaptive Dynamics approach. 
Specifically, after letting consumer and resource abundances reach a steady state, I created a mutant consumer by randomly choosing one and modifying its attack rate on one resource by either subtracting or adding a small constant (0.01 in the following simulations) with equal probability. 
The mutant's attack rate on the other resource was determined by a tradeoff, such that $\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff [@Sargent2006]. 
This function has the useful property that it differentiates between cases where intermediate combinations of $a_{\text{1},j}$ and $a_{\text{2},j}$ are higher than the extremes (when $n>1$, green line in Fig. \ref{fig:tradeoff}) or, conversely, where the two extremes are higher than intermediate investments (when $n<1$, orange line in Fig. \ref{fig:tradeoff}). 
When $n=1$, the tradeoff function is linear, and all combinations of $a_{\text{1},j}$ and $a_{\text{2},j}$ have the same total attack rate (blue line in Fig. \ref{fig:tradeoff}). 
Assuming the mutant consumer was rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. 
If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource abundances to reach a steady state. 
I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy [ESS, @Smith1973] or an evolutionary limit (e.g., $\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}$ is constrained to a maximum of 1 and minimum of 0).

@@

Similarly, the evolutionary tradeoff that favored character displacement decreased resource availability across all foraging scenarios.
Although data on the shape of the tradeoff in consumer foraging traits is scarce, two classic examples of character displacement, Darwin's finches and threespine stickleback, both appear to exhibit a tradeoff where extreme trait values increase the net foraging rate of consumers [@Schluter1985; @Arnegard2014]. 
While it is theoretically possible that character displacement does not alter (or even increase) resource abundances, this was limited to the simplest, and arguably least realistic, foraging scenario and under tradeoffs that did not favor large displacements, and thus less likely to detect in nature.
<!--Taken together, my results call for empirical work to test these clear theoretical predictions and suggest a revision is needed for one of the criteria used to demonstrate character displacement.-->

@@

<!-- Remove for double-blind review
# Acknowledgements

This work was inspired by discussions with Seth Rudman, Dolph Schluter, Ben Gilbert, and Kevin McCann.
Sally Otto provided much help in early analyses of the mathematical model. 
I would not have been able to take the theory as far as I did without her guidance and encouragement.
Matt Osmond, Jean Gibert, and Seth Rudman provided critical feedback on an earlier draft.
For funding support, I thank the University of British Columbia (Four-Year Fellowship to M.A. Barbour), NSERC (Discovery grant to Greg Crutsinger), and the Swiss National Science Foundation (grant 31003A_160671 to Jordi Bascompte).
-->

>>>>

---
output:
  pdf_document:
    keep_tex: true
    toc: false
    number_sections: false
    df_print: paged
    fig_caption: true
    citation_package: natbib 
    #includes:
    #  after_body: Figures_Edited.tex
header-includes:
  - \usepackage{fullpage}
  - \linespread{1.7} 
  - \usepackage{lineno}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
bibliography: References.bib
biblio-style: amnat
fontsize: 11pt
fontfamily: mathpazo
documentclass: article
linkcolor: black
urlcolor: black
citecolor: black
---

@@

@@

*Data accessibility*: No new data were used. All code used to reproduce this manuscript and simulations within are publically available on GitHub (https://github.com/mabarbour/ECD_model) and have been archived with Zenodo (https://zenodo.org/badge/latestdoi/79349192).

@@

where consumer $j$'s feeding rate on resource $i$ is influenced by the abundance of each resource; saturates as resource abundances increase (due to handling time $h_{ij}$); and consumer habitat preferences are modified by the relative abundance of resources, such that: $W_{ij}=w_{ij}R_i/(w_{1,j}R_1+w_{2,j}R_2)$.<!--$W_{ij}=\frac{w_{ij}R_i}{w_{1,j}R_1+w_{2,j}R_2}$.-->

Previous studies have analyzed the evolution of consumer attack rates in the first two foraging scenarios using an Adaptive Dynamics approach, with the general result being divergent character displacement [@Lawlor1976; @Abrams1986]. 
I also used an Adaptive Dynamics approach to analyze the evolution of consumer attack rates in the third foraging scenario, and I too observed divergent character displacement (detailed analysis given in Appendix S1).
I say consumers have undergone divergent character displacement if their evolved attack rates are more specialized when evolving with vs. without a competing consumer. 
Specialization of consumer $j$ on resource $1$ is measured as $a_{1,j}/(a_{1,j}+a_{2,j})$<!--$\frac{a_{1,j}}{a_{1,j}+a_{2,j}}$-->, where a value of 0.5 is a complete generalist ($a_{1,j}=a_{2,j}$), and a value of 1 is a complete specialist ($a_{2,j}=0$).
Values less than 0.5 indicate specialization on the other resource. 
Since I did not observe convergent character displacement in any of the foraging scenarios I analyzed, I refer to divergent character displacement as simply (ecological) character displacement throughout the rest of the text.

@@

To gain insight to the eco-evolutionary feedback generated by character displacement, I conducted simulations using an Adaptive Dynamics approach. 
Specifically, after letting consumer and resource abundances reach a steady state, I created a mutant consumer by randomly choosing one and modifying its attack rate on one resource by either subtracting or adding a small constant (0.01 in the following simulations) with equal probability. 
The mutant's attack rate on the other resource was determined by a tradeoff, such that $\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff [@Sargent2006]. 
This function has the useful property that it differentiates between cases where intermediate combinations of $a_{\text{1},j}$ and $a_{\text{2},j}$ are higher than the extremes (when $n>1$, green line in Fig. \ref{fig:tradeoff}) or, conversely, where the two extremes are higher than intermediate investments (when $n<1$, orange line in Fig. \ref{fig:tradeoff}). 
When $n=1$, the tradeoff function is linear, and all combinations of $a_{\text{1},j}$ and $a_{\text{2},j}$ have the same total attack rate (blue line in Fig. \ref{fig:tradeoff}). 
Assuming the mutant consumer was rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. 
If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource abundances to reach a steady state. 
I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy [ESS, @Smith1973] or an evolutionary limit (e.g., $a_{ij}/(a_{\text{1},j}+a_{\text{2},j})$<!--$\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}$--> is constrained to a maximum of 1 and minimum of 0).

@@

Similarly, the evolutionary tradeoff that favored character displacement decreased resource availability across all foraging scenarios.
Although data on the shape of the tradeoff in consumer foraging traits is scarce, two classic examples of character displacement, Darwin's finches and threespine stickleback, both appear to exhibit a tradeoff where extreme trait values increase the net foraging rate of consumers [@Schluter1985; @Arnegard2014]. 
While it is theoretically possible that character displacement does not alter (or even increase) resource abundances, this was limited to the simplest, and arguably least realistic, foraging scenario and under tradeoffs that did not favor large displacements, and is thus less likely to detect in nature.
<!--Taken together, my results call for empirical work to test these clear theoretical predictions and suggest a revision is needed for one of the criteria used to demonstrate character displacement.-->

@@

# Acknowledgements

This work was inspired by discussions with Seth Rudman, Dolph Schluter, Benjamin Gilbert, and Kevin McCann.
Sarah Otto provided much help in early analyses of the mathematical model. 
I would not have been able to take the theory as far as I did without her guidance and encouragement.
Matthew Osmond, Jean Philippe Gibert, and Seth Rudman provided critical feedback on an earlier draft.
For funding support, I thank the University of British Columbia (Four-Year Fellowship to M.A. Barbour), NSERC (Discovery grant to Greg Crutsinger), and the Swiss National Science Foundation (grant 31003A_160671 to Jordi Bascompte).

### 2020-07-02 manuscript_ECD_model.tex
add author info, acknowledgements, unstack inline equations, and final edit from AE.

@@

where consumer \(j\)'s feeding rate on resource \(i\) is influenced by
the abundance of each resource; saturates as resource abundances
increase (due to handling time \(h_{ij}\)); and consumer habitat
preferences are modified by the relative abundance of resources, such
that: \(W_{ij}=\frac{w_{ij}R_i}{w_{1,j}R_1+w_{2,j}R_2}\).

Previous studies have analyzed the evolution of consumer attack rates in
the first two foraging scenarios using an Adaptive Dynamics approach,
with the general result being divergent character displacement
\citep{Lawlor1976, Abrams1986}. I also used an Adaptive Dynamics
approach to analyze the evolution of consumer attack rates in the third
foraging scenario, and I too observed divergent character displacement
(detailed analysis given in Appendix S1). I say consumers have undergone
divergent character displacement if their evolved attack rates are more
specialized when evolving with vs.~without a competing consumer.
Specialization of consumer \(j\) on resource \(1\) is measured as
\(\frac{a_{1,j}}{a_{1,j}+a_{2,j}}\), where a value of 0.5 is a complete
generalist (\(a_{1,j}=a_{2,j}\)), and a value of 1 is a complete
specialist (\(a_{2,j}=0\)). Values less than 0.5 indicate specialization
on the other resource. Since I did not observe convergent character
displacement in any of the foraging scenarios I analyzed, I refer to
divergent character displacement as simply (ecological) character
displacement throughout the rest of the text.

@@

To gain insight to the eco-evolutionary feedback generated by character
displacement, I conducted simulations using an Adaptive Dynamics
approach. Specifically, after letting consumer and resource abundances
reach a steady state, I created a mutant consumer by randomly choosing
one and modifying its attack rate on one resource by either subtracting
or adding a small constant (0.01 in the following simulations) with
equal probability. The mutant's attack rate on the other resource was
determined by a tradeoff, such that
\(\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1\),
where \(A\) is the total investment in attack rates and \(n\) describes
the shape of the tradeoff \citep{Sargent2006}. This function has the
useful property that it differentiates between cases where intermediate
combinations of \(a_{\text{1},j}\) and \(a_{\text{2},j}\) are higher
than the extremes (when \(n>1\), green line in Fig. \ref{fig:tradeoff})
or, conversely, where the two extremes are higher than intermediate
investments (when \(n<1\), orange line in Fig. \ref{fig:tradeoff}). When
\(n=1\), the tradeoff function is linear, and all combinations of
\(a_{\text{1},j}\) and \(a_{\text{2},j}\) have the same total attack
rate (blue line in Fig. \ref{fig:tradeoff}). Assuming the mutant
consumer was rare, I then determined whether the mutant had higher
relative fitness than the resident consumer, and thus could invade and
replace the resident consumer. If the mutant was able to invade, I
updated the attack rate of the resident consumer to the mutant attack
rate and allowed consumer and resource abundances to reach a steady
state. I then repeated the simulation up to 10,000 times, which was
sufficient for consumers to either reach an evolutionary stable strategy
\citep[ESS,][]{Smith1973} or an evolutionary limit (e.g.,
\(\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}\) is constrained to a
maximum of 1 and minimum of 0).

@@

Similarly, the evolutionary tradeoff that favored character displacement
decreased resource availability across all foraging scenarios. Although
data on the shape of the tradeoff in consumer foraging traits is scarce,
two classic examples of character displacement, Darwin's finches and
threespine stickleback, both appear to exhibit a tradeoff where extreme
trait values increase the net foraging rate of consumers
\citep{Schluter1985, Arnegard2014}. While it is theoretically possible
that character displacement does not alter (or even increase) resource
abundances, this was limited to the simplest, and arguably least
realistic, foraging scenario and under tradeoffs that did not favor
large displacements, and thus less likely to detect in nature.

@@

\bibliography{ReferencesAbbrev.bib}

>>>>

\begin{center} \large Matthew A. Barbour$^{1,2,\ast}$ \normalsize \end{center}

\noindent 1. University of Zurich, Department of Evolutionary Biology
and Environmental Studies, Winterthurerstrasse 190, 8057 Zurich,
Switzerland;

\noindent 2. University of British Columbia, Department of Zoology, 2212
Main Mall, Vancouver, BC V6T 1Z4, Canada;

\(^\ast\) Corresponding author; e-mail:
\href{mailto:matthew.barbour@ieu.uzh.ch}{\nolinkurl{matthew.barbour@ieu.uzh.ch}},
phone: +41 44 635 61 63

\emph{Statement of authorship}: MAB conceived the study, derived
analytical solutions, performed simulations, and wrote the paper.

\emph{Data accessibility}: No new data were used. All code used to
reproduce this manuscript and simulations within are publically
available on GitHub (\url{https://github.com/mabarbour/ECD_model}) and
have been archived with Zenodo
(\url{https://zenodo.org/badge/latestdoi/79349192}).

@@

where consumer \(j\)'s feeding rate on resource \(i\) is influenced by
the abundance of each resource; saturates as resource abundances
increase (due to handling time \(h_{ij}\)); and consumer habitat
preferences are modified by the relative abundance of resources, such
that: \(W_{ij}=w_{ij}R_i/(w_{1,j}R_1+w_{2,j}R_2)\).

Previous studies have analyzed the evolution of consumer attack rates in
the first two foraging scenarios using an Adaptive Dynamics approach,
with the general result being divergent character displacement
\citep{Lawlor1976, Abrams1986}. I also used an Adaptive Dynamics
approach to analyze the evolution of consumer attack rates in the third
foraging scenario, and I too observed divergent character displacement
(detailed analysis given in Appendix S1). I say consumers have undergone
divergent character displacement if their evolved attack rates are more
specialized when evolving with vs.~without a competing consumer.
Specialization of consumer \(j\) on resource \(1\) is measured as
\(a_{1,j}/(a_{1,j}+a_{2,j})\), where a value of 0.5 is a complete
generalist (\(a_{1,j}=a_{2,j}\)), and a value of 1 is a complete
specialist (\(a_{2,j}=0\)). Values less than 0.5 indicate specialization
on the other resource. Since I did not observe convergent character
displacement in any of the foraging scenarios I analyzed, I refer to
divergent character displacement as simply (ecological) character
displacement throughout the rest of the text.

@@

To gain insight to the eco-evolutionary feedback generated by character
displacement, I conducted simulations using an Adaptive Dynamics
approach. Specifically, after letting consumer and resource abundances
reach a steady state, I created a mutant consumer by randomly choosing
one and modifying its attack rate on one resource by either subtracting
or adding a small constant (0.01 in the following simulations) with
equal probability. The mutant's attack rate on the other resource was
determined by a tradeoff, such that
\(\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1\),
where \(A\) is the total investment in attack rates and \(n\) describes
the shape of the tradeoff \citep{Sargent2006}. This function has the
useful property that it differentiates between cases where intermediate
combinations of \(a_{\text{1},j}\) and \(a_{\text{2},j}\) are higher
than the extremes (when \(n>1\), green line in Fig. \ref{fig:tradeoff})
or, conversely, where the two extremes are higher than intermediate
investments (when \(n<1\), orange line in Fig. \ref{fig:tradeoff}). When
\(n=1\), the tradeoff function is linear, and all combinations of
\(a_{\text{1},j}\) and \(a_{\text{2},j}\) have the same total attack
rate (blue line in Fig. \ref{fig:tradeoff}). Assuming the mutant
consumer was rare, I then determined whether the mutant had higher
relative fitness than the resident consumer, and thus could invade and
replace the resident consumer. If the mutant was able to invade, I
updated the attack rate of the resident consumer to the mutant attack
rate and allowed consumer and resource abundances to reach a steady
state. I then repeated the simulation up to 10,000 times, which was
sufficient for consumers to either reach an evolutionary stable strategy
\citep[ESS,][]{Smith1973} or an evolutionary limit (e.g.,
\(a_{ij}/(a_{\text{1},j}+a_{\text{2},j})\) is constrained to a maximum
of 1 and minimum of 0).

@@

Similarly, the evolutionary tradeoff that favored character displacement
decreased resource availability across all foraging scenarios. Although
data on the shape of the tradeoff in consumer foraging traits is scarce,
two classic examples of character displacement, Darwin's finches and
threespine stickleback, both appear to exhibit a tradeoff where extreme
trait values increase the net foraging rate of consumers
\citep{Schluter1985, Arnegard2014}. While it is theoretically possible
that character displacement does not alter (or even increase) resource
abundances, this was limited to the simplest, and arguably least
realistic, foraging scenario and under tradeoffs that did not favor
large displacements, and is thus less likely to detect in nature.

@@

\section{Acknowledgements}\label{acknowledgements}

This work was inspired by discussions with Seth Rudman, Dolph Schluter,
Benjamin Gilbert, and Kevin McCann. Sarah Otto provided much help in
early analyses of the mathematical model. I would not have been able to
take the theory as far as I did without her guidance and encouragement.
Matthew Osmond, Jean Philippe Gibert, and Seth Rudman provided critical
feedback on an earlier draft. For funding support, I thank the
University of British Columbia (Four-Year Fellowship to M.A.~Barbour),
NSERC (Discovery grant to Greg Crutsinger), and the Swiss National
Science Foundation (grant 31003A\_160671 to Jordi Bascompte).

\bibliography{References.bib}
